---
title: "Dendrogram Analysis: Streamwater Chemistry"
author: "Kiera Matiska"
date: "2/15/2022"
output: 
  html_document:
    code_folding: hide
---

## Overview:

This code will analyze stream chemistry data and create a dendrogram for relationships between different sites. The data includes different chemistry metrics, such as ammonium concentrations (uM), nitrate concentrations (uM), phosphate concentrations (uM), total dissolved nitrogen (uM), total dissolved phosphorus (uM), total particulate carbon (uM), total particulate nitrogen (uM), total particulate phosphorous (uM), total suspended solids (mg/L), and specific conductivity (uS/cm). The data set contains 12 sites and 6 of those sites are analyzed in this coding program.

#### Citation:

Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. Environmental Data Initiative. https://doi.org/10.6073/pasta/67a558a24ceed9a0a5bf5e46ab841174. 

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# attach packages
library(tidyverse)
library(here)
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
```

```{r}
stream_chem_read <- read_csv(here("data", "sbc_lter_registered_stream_chemistry.csv")) # read in data

stream_chem_na <- stream_chem_read %>% 
  replace(stream_chem_read == -999.0, NA) # converts -999.0 values to NA

stream_chem <- stream_chem_na %>% 
  drop_na() # drops all na values
```

```{r}
# Subset with mean data by site
stream_chem_site <- stream_chem %>% 
  group_by(site_code) %>% 
  summarize(mean_nh4_uM = mean(nh4_uM), # summarize data by site
            mean_no3_uM = mean(no3_uM),
            mean_po4_uM = mean(po4_uM),
            mean_tdn_uM = mean(tdn_uM),
            mean_tdp_uM = mean(tdp_uM),
            mean_tpc_uM = mean(tpc_uM),
            mean_tpn_uM = mean(tpn_uM),
            mean_tpp_uM = mean(tpp_uM),
            mean_tss_mgperLiter = mean(tss_mgperLiter),
            mean_spec_cond_uSpercm = mean(spec_cond_uSpercm))

# summary(stream_chem_site) 

# scale subset by type
stream_site_scale <- stream_chem_site %>% 
  select(ends_with(c("_uM", "_mgperLiter", "_uSpercm"))) %>% # select only certain variables
  scale() # scale data

# summary(stream_site_scale)
```

## Heirarchical Clustering Analysis: Dendrogram

```{r}
stream_dist <- dist(stream_site_scale, method = "euclidean") 

stream_hc_complete <- hclust(stream_dist, method = "complete") # create logical clusters

stream_hc_complete$labels <- c("AB00", "GV01", "MC00", "MC06", "ON02", "RG01") # changed site names

ggdendrogram(stream_hc_complete, rotate = TRUE) + # defined dendrogram
  theme_minimal() + # changed theme
  labs(x = "Site", y = "Height") # changes axis label
```

**Figure 1.** Dendogram of site relationships. From this dendogram, it can be concluded that sites pair 2 and 5 and pair 3 and 4 are the most related to each other.From pair 3 and 4, branches containing site 6 and 1 indicate lesser relationships and the final branch indicating pair site 2 and 5 are the least related to the other sites.

## Summary:

Overall, we can conclude that:

- sites 2 and 5 and sites 3 and 4 are strongly related to each other
- branch cluster with sites 2 and 5 have poor relationships with branch cluster 3 and 4 due to distance between branches
- sites 1 and 6 are closer related to sites 3 and 4 than to sites 2 and 5
